Privacy Analysis of a Mobile Sensor Application

### 2014-12-18 llncs/paper.tex
Filled in section from LG deliverable.

\newcommand{\TODO}[1]{\textcolor{magenta}{TODO: #1}}

@@

@@

\TODO{later}

@@

\TODO{RG}

@@

\TODO{HH}

@@

\TODO{MM}

@@

@@

>>>>

@@

%% CUSTOM BINDINGS %%

\usepackage{amsmath}

\usepackage{alltt}
\newcommand{\TODO}[1]{\begin{alltt}\textcolor{magenta}{TODO: #1}\end{alltt}}

\newenvironment{LGContent}
{ \par\color{blue} \it \small }
{ \par }

\usepackage{xspace}
\newcommand{\om}{[...]\xspace} % omission

%% END CUSTOM BINDINGS %%

@@

\TODO{(later)}

@@

\TODO{RG\\
* Context: Reference model IT Security analysis\\
* Application: Travel Survey\\
* Different definitions of privacy in the literature.\\
* Study outcomes depend on privacy definition
}

\begin{LGContent}
In recent days the importance of privacy protection has been amplified
by the reports about the mass surveillance of ordinary citizens on a
global scale by the NSA and other intelligence agencies around the
world.

While aiming at the noble cause of enhancing eParticipation using
mobile technologies, Live+Gov systems do process a large variety data
that is potentially infringing the citizens privacy. The captured data
includes personal information like name, phone numbers and email
addresses and sensor data from GPS and accelerometer sensors. Also
with some applications it is possible to gather images and textual
input from the citizen.

While the collection of this data is necessary for providing the
advanced services that Live+Gov aims to deliver, at he same time, the
available raw data can be used to draw a very detailed picture of the
private life of the citizen. For instance can GPS location tracking be
used to reveal shopping habits (e.g. when a car seller is visited) and
associations to political groups (when a meeting is
attended). Accelerometer data can be used to infer medical conditions
like walking disabilities. Images can contain faces of nearby persons
to with whom the citizen is associated. All this data is highly
sensitive to the citizens privacy and can be used against the citizen
if it falls in the wrong hands.

The great importance of protecting the citizens privacy should be
apparent from these examples. The European Union, as well as many
other countries in the past, has set out a number of directives that
regulate the collection, processing and use of privacy sensitive
data. We explain the most relevant legislation in Section
\ref{sec:legal}.

The ethical aspects of privacy have been the subject of study of many
social scientists and philosophers. One scholar which is particularly
relevant in our context is Charles Fried. He investiages, why we are
intuitively so sensitive to violations of our privacy. For him privacy
is not asserted as an intrinsic value by itself, he rather sated:

\begin{quote}
Privacy is not simply an absence of information about us in the minds of others;
rather it is the control we have over information about ourselves.
\end{quote}

Fried`s study on the understanding of privacy provided a great
contribution to the research on the same term in philosophy and
computer science and despite the fact his text was published in 1970,
he already included technologies to its viewpoint (like location
monitoring) that are particularly relevant to our Context.  We explain
his theory in \ref{sec:ethics} and follow his definition of privacy in
this document.

In this document we perform a thorough analysis of how to protect the
privacy of a citizen and present several implementations that form the
core of our privacy aware Sensor Data Storage and Mining Infrastructure.

We identify six main threats for the citizens privacy, and derive
eight recommendations that should be followed in order to reduce the
associated risks for hazards.  This analysis form the guideline for
the selection of 9 measures that were implemented in our system and
documented in Chapter \ref{chap:impl}.
\end{LGContent}

@@

\TODO{HH:\\
* System description: (Raw version copied from Privacy Analysis)\\
* Information needs of transport agency\\
}

\subsubsection{IT-Systems \& Interactions}
\label{subsubsection:it-systems}

\begin{LGContent}
This section outlines the general architecture (Figure \ref{figure:IT Systems}) of IT systems for public monitoring comparable to the Live+Gov project.
This includes a description of it technical infrastructure and the interactions between its components.

The IT infrastructure of the Live+Gov system, i.e. the Live+Gov toolkit and the customization of the software components are described in detail in various project deliverables: D4.1, D4.3, D1.1, D5.1.
In this section we give an abstraction of those systems from the perspective of WP1.

\input{./figures/IT-Systems}

A \emph{citizen} (a) carries a mobile device running the \emph{sensor collector} application (d).
The sensor collector application is able to collect various kinds of sensor data (accelerometer, GPS, GSM, ...) and can sent the raw data back to the \emph{data center} (c).
The sensor collector can also be able to perform certain data mining operations.

Examples of such data mining operations include human activity recognition,the detection of service lines, detection of characters and faces from images.

In the Live+Gov Project mobile devices are used in particular for the following activities:
\begin{enumerate}
\item collection of GPS samples,
\item mining of Human activities (HAR) based on accelerometer samples,
\item collection of reports consisting of an image, free text and selected categories.
\end{enumerate}

The \emph{data center} stores and processes sensor data collected with the sensor collector application. It can also take into account data obtained from third parties, like the current positions of trains.
The data center can sent mining end products (traffic jam reports, bus schedule) back to the mobile device of the citizen.

In the Live+Gov Project data centers, in particular, perform the following activities:
\begin{enumerate}
\item storage of login credentials, name, and email address for each citizen,
\item storage of collected GPS samples (user id, timestamp, GPS location),
\item storage of HAR results (user id, timestamp, HAR result),
\item detection of service lines based received GPS samples (SLD),
\item storage of SLD results (userid, timestamp, SLD result),
\item sent back SLD results to mobile device,
\item storage of received reports (user id, timestamp, report),
\item detection of inherent patterns of the received reports.
\end{enumerate}

The \emph{System Provider} (b) provides and operates technical infrastructure like the Data Center and the \emph{Reporting Tool} (f).
The Reporting Tool queries the Data Center for aggregated data to visualize in form of charts and other means suitable to help understanding of monitored citizens.

\textit{Local Authorities} (c) use the reporting tool to get information in order to understand citizen movement and improve public services.
In such systems, the most valuable information for local authorities is not the raw sensor data, but aggregated views on the mining end products.

In the Live+Gov Project the reporting tools allows, in particular the following queries:
\begin{enumerate}
\item show aggregate information about which routes citizens take starting from a given location,
\item show the average waiting time of a citizen for each bus stop,
\item show routes where citizens where running to catch a bus,
\item show locations of all reports in a given time window,
\item visualize detected patterns in the report data,
\item view individual reports.
\end{enumerate}
\end{LGContent}

@@

\TODO{MM:\\
- Shorten\\
- Remove Implicit Type Violations\\
- Update figure\\
- RG: Discuss alternative definitions of privacy\\
}

\begin{LGContent}
Defining privacy is a challenge which seems impossible. This is well put to words by Serge Gutwirth, who notes:

\begin{quote}
The notion of privacy remains out of the grasp of every academic chasing it. Even when it is cornered by such additional modifiers as `our' privacy, it still finds a way to remain elusive. \cite{Gutwirth}
\end{quote}

%Many researchers seem to only ``focus on the ways in which privacy can be infringed'' \cite{7ToP}.
%Thus they try to create taxonomies of \emph{privacy harms} instead of taxonomies of \emph{privacy types}.
%Those two differ in the respect that the former focuses on threats to prohibit whereas the latter focuses on values to protect.
%So one should rather evaluate what aspects are precious about privacy and develop measures to ensure their security than only forbid single actions against it \cite{7ToP}.

Many researchers seem to only ``focus on the ways in which privacy can be infringed'' \cite{7ToP}.
Thus they invest a great amount of work in defining threats to prohibit instead of describing why privacy is so valuable to us.
This way around one could derive measures to ensure and secure its value \cite{7ToP}.
One particular scholar who does this in the context of digital monitoring is Charles Fried.
He questioned, why we are intuitively sensitive to violations of privacy.
But he did not assert privacy as an intrinsic value by itself, he rather sated:

\begin{quote}
Privacy is not simply an absence of information about us in the minds of others;
rather it is \textbf{the control we have over information about ourselves}. \cite{CFried:Privacy}
\end{quote}

According to Fried privacy is a \textit{rational context}.
Which means that one is aware of such context's existence during a rational action.
If we are to share private information with others, we are most likely aware that this action is privacy related.
In that case we are able to selectively disclose information along the two dimensions of quantity and quality.
This leads Fried to his thesis, that privacy is one's ability to create and modulate his or hers social relationships, namely: friendship, love and trust. \cite{CFried:Privacy}

Depending on the conversation partner, we change the degree of intimate information we share if it is a total stranger, colleague, close friend or lover.
With close friends or lovers we share information of great intimacy we do not share with anyone else.
Moreover, we trust those persons to not reveal information about us to others by respecting their privacy.
Trust needs the possibility of unknown failure.
If we would constantly monitor our partners, they cannot fail unnoticed nor can they willingly share that information with us.
Thus they could not trust us anymore. \cite{CFried:Privacy}

So privacy or the its possibility, according to Fried, is the foundation of our core relations: friendship, love and trust.
And thus it is valuable, because those relations are essential to human society \cite{CFried:Privacy}.
Using his anatomy of privacy as foundation of our analysis is suitable because of two points.
At first, Fried's study on the understanding of privacy provided a great contribution to the research on the same term in philosophy \cite{sep-privacy} and computer science \cite{SociotechnicalArchitectureForOnlinePrivacy}.
Secondly, despite the fact his text was published in 1970, he already included technologies to its viewpoint that did not only monitor location, but also record biometric data.

\subsection{The Seven Types of Privacy}\label{sec:privacytypes}

In Fried's definition of privacy as control over "information about onself'', the specification of what constitues such information remains open.
There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis.
To this end we use the the categorization by Friedewald, Finn and Wright \cite{7ToP} called the \textit{Seven Types of Privacy}.
The seven types of privacy are an extension to the four types of privacy by Roger Clarke \cite{RClarke:4ToP}, which are:
\begin{itemize}
\item Privacy of the Person
\item Privacy of Personal Behaviour
\item Privacy of Personal Communication
\item Privacy of Personal Data
\end{itemize}
It is important to note, that this categories do not form no taxonomy, since the categories are not mutually exclusive.
For instance a written email is considered personal communication as well as personal data stored on a computer.

Moreover, Friedewald et al. argue that Clarke's taxonomy is outdated and no longer adequate in order to describe the privacy aspect of our modern, technology driven, world.
In order to address this shortcoming they extend the former four to the now introduced seven types privacy as follows:

\begin{enumerate}

\item \textbf{Privacy of the Person}
This privacy type is generally concerned with one could best understand as biometric privacy.
Friedewald et al. paraphrase it as \emph{``\om the right to keep body functions and body characteristics \om private''}.
This includes but is not limited to measures like weight, height or shoulder width;
biometric identifiers like fingerprints and DNA sequences;
or medical conditions such as limping or having a cold.

\item \textbf{Privacy of Behaviour and Action}
This privacy type is concerned with one's activities in public as well as in private spaces.
It includes but is not limited to religious practices, political activities and sexual preferences or habits.

\item \textbf{Privacy of Communication}
This privacy type is concerned with one's communication in a broad sense.
It includes written correspondence, but also conversations conducted either vis-a-vis or via electronic devices.
Friedewald et al. put it as the right to free discussion without unknown interception by third parties.

\item \textbf{Privacy of Data and Image}
This privacy type is concerned with the secrecy of personal data, especially its automatic disclosure to other individuals and organizations.
It includes data such as paychecks, insurance information or records of public administration.
However, it also refers to pictures taken without consent and digital identifiers like IP addresses or social security numbers.

\item \textbf{Privacy of Thoughts and Feelings}
This privacy type is the counterpart to Privacy of the Person like body and mind are counterparts of one another.
Comparable to the Privacy of Data and Image, Friedewald et al. state that one's thoughts and feelings must not be automatically revealed to others.
This could simply happen by the disclosure of one's diary or by technologies which allow emotion detection through biometric means.
One's body temperature or iris reflexes might infer stress or excitation.

\item \textbf{Privacy of Location and Space}
This privacy type is concerned with one's movements in public spaces and the protection of private spaces.
Friedewald et al. qualify the first dimension as one's right to move without being identified, tracked or monitored.
The second dimension is qualified as one's general right to solitude, especially the right to the inviolability of the home.

\item \textbf{Privacy of Association}
This privacy type is also put as group privacy.
Friedewald et al. state that one must have the possibility with whomever without being recorded.
Associations like friends or organizations such as political parties must not automatically be recorded because one associates with them, and vice-versa.
\end{enumerate}

\subsection{Privacy Type Inferences}

The above types of privacy types are not disjoint to each other.
For example if the location of a citizen is known it is possible to infer information about political associations (e.g. his visits to a party meeting).
In this section we explore possibilities how personal data can be inferred from each other.
Our findings are summarized in Figure \ref{figure:Implicit Privacy Violation Matrix}.

%\includegraphics{../diagrams/png/implicit-privacy-violation-matrix.png}
\input{./figures/Implicit-Privacy-Violation-Matrix}

\paragraph*{1. Privacy of The Person}

The Privacy of The Person is concerned with one's biometric privacy.
If this type is violated, following implicit violations are possible:

\begin{itemize}
\item [(1-5)]
  Privacy of Thoughts and Feelings. Some psychological diseases
  (e.g.~depression) have physiological impact. Such physiological
  patterns could be detected.
\end{itemize}

\paragraph*{2. Privacy of Behaviour and Action}

The Privacy of Behaviour and Action is concerned with one's social, political, religious, sexual, etc. activities.
If his type is violated, following implicit violations are possible:

\begin{itemize}
\item [(2-1)] Privacy of The Person.
  Religious practices which include body modifications (e.g.~circumcision).
\item [(2-5)] Privacy of Thoughts and Feelings.
  Social activities in general depend on a certain intellectual attitude.
  Such an activity is the expressions of such an attitude.
\item [(2-7)] Privacy of Association.
  Recording religious, political or sexual activities can reveal association with churches, political parties or sexual partners.
\end{itemize}

\paragraph*{3. Privacy of Communication}

The Privacy of Communication is concerned with not having such communication (correspondence or vis-a-vis) intercepted.
This is very broad type of privacy.
Depending on the contents of the intercepted communication every other type can be violated:

\begin{itemize}
\item [(3-1)] Privacy of The Person. Communication about body characteristics.
\item [(3-2)] Privacy of Behaviour and Action. Communication about social activities.
\item [(3-4)] Privacy of Data and Image.
  Communication containing one's passwords or other sensitive data.
\item [(3-5)] Privacy of Thoughts and Feelings.
  Communication of thoughts and feelings, e.g.~wiretapping a flirt or a catholic confession
  ritual.
\item [(3-6)] Privacy of Location and Space.
  Interception of face-to-face communication is only possible if one's location and space is violated (wiretapping).
\item [(3-7)] Privacy of Association.
  Communication about one's associations (family members, churches, etc.).
\end{itemize}

\paragraph*{4. Privacy of Data and Image}

The Privacy of Data and Image is concerned with one's data not being automatically available to others.
This also is a very broad type of privacy.
Depending of the data or image contents every other type can be violated:

\begin{itemize}

\item [(4-1)] Privacy of The Person.
  Images or stored biometric information reveal one's physical characteristics.
\item [(4-2)]Privacy of Behaviour and Action.
  Images or diaries can reveal one's social activities.
\item [(4-3)]Privacy of Communication.
  Modern communication systems usually contain some sort of archive function, e.g.~E-mail clients do not automatically delete messages.
  Such messages are data and reveal one's communication.
\item [(4-4)]Privacy of Thoughts and Feelings.
  Images can show one's emotional state.
\item [(4-5)]Privacy of Location and Space.
  Images can reveal one's location, e.g.~making a picture in front of the Eifel Tower.
\item [(4-6)]Privacy of Association. E-mail data can also reveal
  association.
\end{itemize}

\paragraph*{5. Privacy of Thoughts and Feelings}

The Privacy of Thoughts and Feelings is concerned with keeping such
thoughts and feelings secret. If this type is violated, following
implicit violations are possible:

\begin{itemize}
\item [(5-1)] Privacy of The Person.
  Thoughts and feelings can reveal medical conditions.
\item [(5-2)] Privacy of Behaviour and Action.
  Thoughts and feelings can reveal a certain attitudes which create a foundation for certain   social activities.
\item [(5-7)] Privacy of Association.
  Thoughts and feelings can reveal individual association, e.g amorous feelings for a certain person.
\end{itemize}

\paragraph*{6. Privacy of Location and Space}

The Privacy of Location and Space is concerned with one's right to move freely without being tracked and one's right to private places.
If this type is violated, following implicit violations are possible:

\begin{itemize}
\item [(6-1)] Privacy of The Person.
  Frequently visited doctors can reveal certain medical conditions, if such doctors are known specialsts.
  In general it could imply illness.
\item [(6-2)] Privacy of Behaviour and Action.
  Frequently visited places in general can reveal association and hence implies social activities.
\item [(6-3)] Privacy of Communication.
  If one's location is knwon, it is possible to intercept (wiretap) one's communication.
  This also may violate the right to private spaces.
\item [(6-4)] Privacy of Data and Image.
  If one's location is known, it is possible shoot pictures.
\item [(6-5)] Privacy of Thoughts and Feelings.
  Frequently visited persons may imply certain thoughts and feelings, e.g.~having a mistress.
\item [(6-6)] Privacy of Association.
  Frequently visited places can reveal associations simply by searching in maps or yellow-pages.
\end{itemize}

\paragraph*{7. Privacy of Association}

The Privacy of Association is concerned with one's right to associate with whomever one wants, without that association having recorded.
If this type is violated, following implicit violations are possible:

\begin{itemize}

\item [(7-1)] Privacy of The Person.
  Association with tattoo artists could imply having tattoos or other body modifications.
\item [(7-2)] Privacy of Behaviour and Action.
  Association with churches or political organizations could imply certain activities.
\item [(7-5)] Privacy of Thoughts and Feelings.
  Association with churches of political organizations could imply a certain intellectual attitude.
\end{itemize}

\end{LGContent}

\subsection{Sensor Data Privacy Impact}\label{sec:SensorPrivacyImpact}

\begin{LGContent}

\input{./figures/LG-Implicit-Sensor-Privacy-Matrix}

%TODO: Reformulate the section.
% - Too much overlap with preceeding sections.
% - Structure should be per Sensor Type
% - Add HAR and SLD!

In this section we analyze the impact of the disclosure sensor data and certain processing results to the citizens privacy.
This analyisis builds upon the preceeding analyis, but is more focused on the concrete type of data vailable.
For example, the disclosure of Service Line Detection results, does violate the Privacy of Location and Space but not the Privacy of Association that is violated by general GPS tracking.

\textbf{GPS Data}.
The GPS sensor gives the current longitude and latitude, the current global position of the mobile device and its carrier, although there is some artificial inaccuracy within civil use. Threefore, the collection of GPS data violates directly the citizens privacy of Location and Space, and the privacy of Data an Image.

By inference we also get implicit violations of the Privacy of the Person, Privacy of Behavior and Action and Privacy of Association.

\textbf{Motion Sensors.}
Accelerometer, Rotation Vector, Gyroscope and Magnetic field sensor measure the physical movement of the mobile device on all three axes.
If the mobile device is carried ``normally'' its safe to say that those sensors also measure the moments of its carrier.
So his privacy is infringed regarding biometric behaviour, as it is captured automatically
Privacy of Data and Image is trivially threatened because here sensor data is individual data, a priori.

By inference we also get implicit violations of the Privacy of Behavior and Action.

\textbf{Network Sensors.}
The GSM and WLAN sensors reveal the position of the mobile device and its carrier, when used in connection with external databases.
The GSM sensor gives the exact cell, the mobile device has registered with at the current moment.

The Bluetooth sensors record lists of the bluetooth clients in the direct neighbourhood.
Since those clients are usually moving, inference of the position is usually not possible.
Instead, bluetooth clients carried by a third person may infringe the Privacy of Association.

A similar argumentat applies to WLAN snsors. If one frequently connects with an organizational wireless network, e.g.~an university network, an association can be deduced (student or staff).

\textbf{Human Activity Recognition.}
The detection and collection of human activities like walking, standing and running, can interfer with the Privacy of the Person, e.g.~since these movement patterns can be indicators for a person's health. Also, trivially, Privacy of Data and Image is violated.

\textbf{Service Line Detection.}
The detection and collection of the service line the user currently uses in the public transportation system allows inference of the location of the user, at least when entering and leaving those service lines at e.g.~bus stops.
Implicit privacy violations apply accordingly.

\end{LGContent}

@@

\section{IT Security Reference Model}

\begin{LGContent}

The goal of this chapter to analyze and identify the threads to
personal privacy that are posed by collecting, storing and processing
sensor data from mobile phones.  We derive concrete privacy protection
measures that address the main risks involved with handling such data.

The complexity of our systems and the variety of threads make a great
number of counter measures plausible.  We approach this complexity
with the aid of a general security analysis model developed in
\cite{Grimm:ItSecRefModel}.  We give a brief introduction to this
model and perform a IT Security Analysis with respect to the privacy
asset for our system.

We follow the Reference Model for IT Security Analysis as described in
\cite{Grimm:ItSecRefModel}.  It supersedes earlier efforts by
e.g. \cite{Avizienis}.  The reference model consists of a \emph{model}
and a \emph{procedure}.  The model aims to organize common security
terminology in a reasonable and practical way.  The procedure
describes a method to analysis the IT system based on that model.  In
this section we give a brief overview over the reference model.

\subsection{Model}

%\includegraphics{../diagrams/png/itsec-ref-model-grimm.png}
\input{./figures/The-IT-Security-Reference-Model-Grimm}

The model is depicted in Figure \ref{fig:refmodel}.  It is organized
in four views (round boxes) that contain a number of components
(rectangular boxes).

%%%%%%%%%%%%%%%%%%%%

The \emph{world view} contains all components describing the current state.
It consists of the following components:
\begin{itemize}
\item \textbf{Actors.}
All identifiable stakeholders of the system under study.
Typical actors include, users, developers, clients and externals.

\item
\textbf{Assets.} Things of value to one or more stakeholders.
The value can be hard (money, data, etc.) or soft (trust, privacy,etc.).
In our case the only asset we are concerned with is the privacy of the citizen.

\item \textbf{IT-Systems.}
The relevant IT-Systems under study. This encompasses hardware (e.g. servers, network infrastructure), as well as software and third party services.
The level of granularity has to be detailed enough to express all possible threats to the assets at stake.

\item \textbf{Conflicts of Interests.}
Different actors have different interests which can be in conflict which each other.
These conflicts of interest are the origin of all attacks to the system.

A typical conflict is the \emph{Criminal-User-Conflict}:
A user wants to keep control over their private data.
A criminal wants to gain money.
The possibility of selling private data (user profiles) to advertisers, renders both interests conflicting.

\item \textbf{Vulnerabilities.}
All identifiable weaknesses in the IT-System.

In the example of the criminal-user-conflict, the criminal has to exploit a vulnerability, e.g. a weak password, to gain access to the private data about the user.

\item \textbf{Interactions.}
This point captures all possible interactions between assets, IT-Systems, humans and vulnerabilites.
It is described in more detail in the next view.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%

The \emph{potential view} displays the intended and unintended interactions of the components in the world view.
The intended interactions support the underlying business objectives.
Unintended interactions lead to threats.
The potential view consists of the following components.
\begin{itemize}
\item \textbf{Business Objectives.}
Interaction of IT-system and actors that realize a business goals of the system owner.

\item \textbf{Threats.}
A threat is a potential interaction that destroys or harms assets of the system.
Concrete realizations of threats can be \emph{attacks} or \emph{accidents}.
Attacks are executed by an actor in response to a conflict of interest.
Accidents are harmful interactions that are not willfully caused by an actor.

\item \textbf{Chances/Risk.}
Evaluation of chances and risks associated to the business objectives and threats.
The risk associated to a threat is its expected loss.
A chance associated to a intended interaction is its expected gain.

In the case that, the loss can be quantified monetary, and the likelihood of occurrence of a threat can
be modeled probabilistically, the risk is given by the product
\[ \text{risk} = \text{loss} \cdot P[\text{threat}]. \]
In practice such a quantitative risk evaluation is often not possible, and a qualitative, heuristic, analysis is performed instead.

\item \textbf{Security Requirements.}
A set of interactions (e.g. threats) that shall not occur within the system in order to achieve its business objectives.
Security requirements are targeted to protect one or more assets.

An example of a security requirement is that a given communication channel shall not be infringed by externals.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%

The \emph{concept view} is a realization of the potential view of the system.
It specifies important interactions that require further planning.
It contains the following components:
\begin{itemize}
\item \textbf{Business Model.}
The plan to achieve business objectives.

\item \textbf{Accident Scenario.}
A concrete outline of an interaction that leads to an accident.
In particular the asset under threatened and exploited vulnerability need to be described.

\item \textbf{Attack Technique.}
A specific technique or technology to attack IT-Systems (Man in the Middle, Phishing, etc.).
In particular the attacking actor, the conflict of interest and the exploited vulnerability need to be described.

\item \textbf{Security Measures.}
It describes a plan of sufficient measures to secure the intended interactions and to avoid the unintended interactions.
Each security measure targets a vulnerability of the system in order to reduce a risk for a certain thread.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%

The \emph{event view} contains all actual events through out the lifetime of the system.
The event view instantiates the concept view of the system.
It contains the following components:
\begin{itemize}
\item \textbf{Business Process.}
The actual, running instance of the business model.

\item \textbf{Accidents.}
All actually happened accidents.

\item \textbf{Attacks.}
All actually happened attacks.

\item \textbf{Security Operations.}
Instances of security measures.
\end{itemize}

\subsection{Procedure}

The analysis procedure is an incremental and iterative process following the four views of the previously described model.

%\includegraphics{../diagrams/png/itsec-ref-model-grimm-procedure.png}
\input{./figures/The-IT-Security-Reference-Analysis-Grimm}

\paragraph*{Step 1. World Analysis}

At first, one has to outline the current state of the system under study. This includes description of:
\begin{itemize}
\item all \textbf{Assets} which must be protected
\item all relevant \textbf{IT-Systems}
\item all involved \textbf{Humans} and their \textbf{Conflicts of Interests}
\item all known \textbf{Vulnerabilities}
\item and all important \textbf{Interactions} between the former components.
\end{itemize}

\paragraph*{Step 2. Potential Analysis}

Secondly, one needs to outline the potential interactions of the system under study.
This includes both the unintended interactions (threads) and the intended interactions (business objectives).
This step produces four artifacts:
\begin{itemize}
\item a \emph{threat specification}, which identifies the threat, its targeted assets, the involved actors and their conflicts of interest.
\item a \emph{threat risk evaluation}, which quantifies the likelihood of a threat manifestation in relation to its associated loss.
\item a \emph{security requirement specification}, which specifies requirements in order to deal with identified hazards
\end{itemize}

\paragraph*{Step 3. Concept Development}

Based on \textbf{Step 2.}, the identified hazards are used alongside realistic accident scenarios and attack techniques to create a \emph{risk matrix}.
With this matrix it is possible to decide if the risk is acceptable or not.
Together with the previously specified security requirements, the matrix is used to define adequate security measures.
Like a business model is an abstract concept to achieve business objectives, this step creates an concept to improve the system's security.

\paragraph*{Step 4. Concept Deployment}

Finally, the security measures have to be implemented.
Additionally, all business operations, accidents, attacks and executed security operations will be recorded in the following time.

The implementation of security measures changes the world view (e.g. IT-systems, actors) and renders the conducted analysis outdated.
So this analysis procedure needs to be conducted again.

\subsection{Abstraction Levels of the Reference Model}

The Reference Model can be used on different levels of abstraction.
This means each component can be used within a wide range of granularity, for instance the security measure \emph{Encryption} can be explored in general or on the level of different concrete encryption tools; or on the even finer level of concrete algorithms.

The utilized abstraction level is not important for the analysis procedure, it depends on the intended audience for the analysis.
However, it is important to use one abstraction level consistently through out the analysis.
\end{LGContent}

@@

\begin{LGContent}
\subsection{Step 1. World Analysis}

\subsubsection{Assets: Privacy}

In this document we focus our attention to only one asset: The privacy of the citizen.

%TODO: UPDATE
Our definition of privacy is described in detail in Chapter \ref{chap:privacy}. In Section \ref{sec:taxonomy} we define privacy as the ``control over private data'' and introduce the following seven different types of privacy:
\begin{enumerate}
\item Privacy of the Person
\item Privacy of Behaviour and Action
\item Privacy of Communication
\item Privacy of Data and Image
\item Privacy of Thoughts and Feelings
\item Privacy of Location and Space
\item Privacy os Association
\end{enumerate}

We refer to Section \ref{sec:taxonomy} for more details.

%
%
% SECTION: IT SYSTEM DESCRIPTION MOVED ABOVE
%
%

\subsubsection{Actors}
\label{subsubsection:humans}

This section describes the human actors previously introduced within the IT system architecture (\ref{subsubsection:it-systems}).
Also the additional actor \textit{External} is introduced as a person with no special access rights.

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams, that help them to avoid long waiting times.
Also they generally benefit from improvements of public infrastructure by local authorities, which is triggered by issue reports.

By using the application citizens are sharing personal information like name and address, as well as data gathered from mobile sensor with the service provider.
This data can be exploited in ways that are harmful to the citizen.
This need to protect this protection is manifest in several laws and constitutions as the right to privacy and data protection.
The citizen has a vital interest in having his legal rights protected and enforced.

If the right to privacy is violated, there is a magnitude of potential harms that interfere with other interests of the citizen.
This includes just annoying spam, where their technical identity is used to send unwanted commercials.
More severe phishing attacks can exploit personal information, and try to manipulate citizens into disclosing credentials like TAN numbers and disrespect their financial interests.
Information about the current location (GPS) or frequently used routes (stalking) can be used to attack and directly harm the health of the citizen.
Information about medical conditions inferred by sensor data (e.g. fintness trackers) are of interest to insurance companies, which may affect the pricing of policies and can interfere with financial interests of the citizen.

Also it is well established (cf. \cite{GuardienMassSurveillance}) that the very act of being monitored can have impact on mental health and performance, promotes distrust and breeds conformity.

In short, citizens are interested in
\begin{itemize}
\item physical wellbeing and health
\item financial profit
\item convenience
\item legitimate use of personal data
\item non-disclosure of personal data to peers of the citizen
\item not being monitored.
\end{itemize}

\textbf{External.}
Externals are persons who do not have privileged access to the IT systems, and are willing to break laws, security constrains and norms in order to promote their interests.

If the external is in some kind of relationship to the citizen, like a friendship or business partnership, the external can have a direct interest in gaining information about the citizen in order to increase their power.

Another common interest of an external is financial profit. For example they want to obtain access to critical systems to steal sensitive data or to get the system under their control.
Controlled systems could be leased as part of a bot net.
Stolen data could simply be sold as is or used for illegitimate purposes, e.g. spam or phishing attacks - or excessive data mining.

Because local authorities are involved in the general outline of the Live+Gov system, the possibility for politically motivated attacks is given.

Externals could want to harm or destroy the systems in order to damage the reputation of local authorities (politicians or other officials) or to make a political statement of their own.

Another possible motivation for external  activities could be social appreciation.
A hacker could attack critical infrastructure just to prove his skills.

In short, externals are interested in
\begin{itemize}
\item increase power over citizen
\item financial profit
\item political activism
\item social standing.
\end{itemize}

\textbf{System Provider.}
System Providers operate the technical infrastructure (hardware and software) of the IT System.
They are private companies and legal persons in their own right, but also employ a number of people with diverging interests, including:
administrators, who maintain and operate the running system;
programmer/developer, who develop the system;
a support manager, who handles customer relations.

As companies, they are interested in gaining financial profit.
Among other things, this depends on customer satisfaction, employee happiness and task complexity.
Unsatisfied customers may not want to pay for the service or do not continue the business relation.
Moreover, unsatisfied customers can create a bad reputation, which affects the market for future customers.

Customer satisfaction is connected with the quality of the offered product or service.
This quality depends on the happiness of employees.
Employees have a claim to professional excellence.
They want to deliver a good job within their means.
If employees cannot satisfy their demand for professional excellence, they might get discontent and deliver poor work.
Moreover, unhappy employees can produce higher costs through sick days.
The worst case scenario could be, that a discontent employee gets angry and steals data or harms the running systems.

At last, the financial success of system providers depends on the task complexity of the maintained infrastructure.
The complexity of a task has to be in reasonable bounds, so that system providers can complete it within time, with a satisfying quality.
If a task has a higher complexity than expected, financial loss is almost certain.
Either System Providers need to hire additional competence to meet schedule and requirements.
Or system providers they stress the time-line, which also results in a higher man-hour salary ratio and additionally endangers customer satisfaction.
Ultimately, high task complexities can affect employee happiness, if employees cannot complete it within their claim to professional excellence.

In short, System Providers are interested in
\begin{itemize}
\item financial profit
\item good working conditions
\item professional excellence
\item manageable complexity.
\end{itemize}

\textbf{Local Authority.}
Local authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers.
They purchase a system specialized for their needs.
For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow.

Such systems are investments, and so naturally local authorities are interested in a profitable return, like increased ticket sales.
However, the return of investment is not directly of financial nature.
Like service providers their financial gain depends on customer satisfaction.
Customers for local authorities are either citizens, who use their services, or politicians, who order their services.
The satisfaction of both sides is interdependent.

Citizens are satisfied customers if the services, e.g. public mobility, work well.
If citizens are happy, it is more likely that politicians gain reputation, as they organize the public services through local authorities.

The first important step of improving the public services is by obtaining business intelligence.
For the urban mobility scenario the Live+Gov systems provide insight in form of traffic jam detection and usage pattern mining, which allow local authorities to focus their efforts to the most important sites.

Additionally, since local authorities act like corporations comparable to service providers, they are also interested in good working conditions.
Discontent employees may harm the system by e.g. disclosure of privacy sensitive data.

In short, local authorities are interested in
\begin{itemize}
\item financial profit
\item political reputation
\item business intelligence
\item good working conditions.
\end{itemize}

\subsubsection{Conflicts of Interests}
\label{subsubsection:Conflicts of Interests}
This section outlines the Conflicts of Interests (Figure \ref{figure:Live+Gov Conflicts of Interests}) between the actors of the proposed IT system architecture.

The individual interests of all actors is already described in the previous section and are not elaborated any further.
The emphasis here is put on prominent existing conflicts, because they provide a foundation for vulnerabilities and subsequent threats.

\input{./figures/LG-Conflicts-Of-Interests}
%TODO: Adapt Figure

\textbf{System Complexity vs Privacy.}
System Providers offer a service to Local Authorities, which is to provide and maintain a monitoring and mining system, e.g. for public mobility.
This system shall produce business intelligence, so that Local Authorities can improve their public services.
This task in it self has a high technical complexity and is the sole asset with financial return for System Providers.
However, this task operates on privacy sensitive data provided by monitored Citizens.
In order to ensure their privacy, System Provider would have to implement additional mechanisms, which allow Citizens to exercise control of their data.
This will not only raise the complexity of the monitoring and mining system, System Providers also have to layout the complexity in a comprehensible manner.
To effectively enable Citizens to preserve their privacy, they need to know what happens with their data.

\textbf{Business Intelligence vs Privacy.}
Local authorities order a monitoring and mining system from system providers, which allows them to produce business intelligence for public services.
The system is an investment for local authorities, so they are interested in as much intelligence as possible to achieve a profitable return.

The gained intelligence is the result of data mining conducted on privacy sensitive data of participating citizens.
They are interested in the successful usage of their data, in a sense that they are also benefactors, e.g. improvement of public mobility.
The main interest of citizens lies in maintaining control over their data and protecting their rights to privacy.
In order do that, they need full disclosure of the processing steps and the purposes their data is used for, and to be given a choice whether such processing should be allowed for their own data.
